---
title: "Peer Assessment 1"
author: "Francis A. Méndez Mediavilla"
date: "December 10, 2014"
output: html_document
---

``` {r loadLibraries, echo=FALSE}
library(lattice)
library(Hmisc)
options(scipen=10)
```

## Loading and preprocessing the data

The number of steps taken were collected for the period of October 1, 2012 until November 30, 2012.

```{r readFile}
dat<-read.csv(file="activity.csv", header=T)
head(dat)
tail(dat)
```

## What is mean total number of steps taken per day?

Compute the total number of steps taken per day.  The first step is to aggregate steps per day.  `steps` is defined as the number of steps taking in a 5-minute interval.  

``` {r totNumSteps}
steps.agg<-aggregate(steps~date, 
                     FUN=sum,
                     data=dat)
```


```{r histSteps, echo=TRUE}
histogram(steps.agg$steps,
          type="density",
          xlab="Number of Steps",
          main="Histogram for the Total Number of Steps per Day")
```

## Calculate and report the mean and median total number of steps taken per day

``` {r MeanMedian}
steps.mean.agg<-mean(steps.agg$steps)
steps.median.agg<-median(steps.agg$steps)
```

The mean total number of steps per day is `r round(steps.mean.agg,0)` and the median number of total steps per day is `r round(steps.median.agg,0)`.

## What is the average daily activity pattern?

### Imputation of missing values.

```{r MissingVals}
num.missing<-sum(is.na(dat$steps))
```

The total number of missing values in the dataset is `r num.missing`.

### Impute the missing values
```{r imputation}
# Compute the mean number of steps by interval
steps.mean<-aggregate(steps~interval, 
                     FUN=mean,
                     data=dat,
                     simplify=T)
# Create variable "dat$steps.imp2".  Contains imputed values.
dat$steps.imp2<-numeric(length(dat$steps))
for (j in 1:length(steps.mean)){
        for (i in 1:length(dat$steps)){
                if(dat$interval[i]==steps.mean$interval[j]){
                        dat$steps.imp2[i]<-steps.mean$steps[j]
                }
        }
}
# Replaces NAs with imputed values
for (i in 1:length(dat$steps)){
        if(is.na(dat$steps[i])){
                dat$steps[i]<-dat$steps.imp2[i]
        }
}
# Deletes the variable containing imputed values
dat$steps.imp2<-NULL

```

## What is the average daily activity pattern?

A time series depicts the behavior of the average daily pattern for the number of steps by 5-minute intervals.

```{r timeSeries}
# Recompute the means with the imputations in place
steps.mean<-aggregate(steps~interval, 
                     FUN=mean,
                     data=dat,
                     simplify=T)

plot(steps.mean$interval,
     steps.mean$steps, 
     type="l", lwd=2,
     xlab="5 Minutes Interval", 
     ylab="Average Daily Steps",
     main="Average Daily Steps")

```

### Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r max num steps}
max.interval<-steps.mean[which.max(steps.mean$steps),]$interval
max.steps<-steps.mean[which.max(steps.mean$steps),]$steps
```

The maximum number of average daily steps (i.e., `r max.steps`) occurs at interval `r max.interval`.

## Are there differences in activity patterns between weekdays and weekends?

```{r flagWeekDays}
# If the day is a weekday, variable WeekDay is set to TRUE
dat$WeekDay<-(weekdays(as.POSIXct(strptime(dat$date, "%Y-%m-%d")))!="Sunday" & weekdays(as.POSIXct(strptime(dat$date, "%Y-%m-%d")))!="Saturday")

weekday.steps.mean<-aggregate(steps[dat$WeekDay==TRUE]~interval[dat$WeekDay==TRUE], 
                     FUN=mean,
                     data=dat,
                     simplify=T)

weekend.steps.mean<-aggregate(steps[dat$WeekDay==FALSE]~interval[dat$WeekDay==FALSE], 
                     FUN=mean,
                     data=dat,
                     simplify=T)

mean.steps.weekday<-round(mean(weekday.steps.mean[,2]),2)
mean.steps.weekend<-round(mean(weekend.steps.mean[,2]),2)

```
The mean number of average steps for the weekdays is `r mean.steps.weekday` steps.

The mean number of average steps for the weekends is `r mean.steps.weekend` steps.

```{r TSPlots, fig.align='center'}
d1<-cbind(weekend.steps.mean$steps,weekend.steps.mean$interval)
d2<-data.frame(d1,WeekDay="weekend")
d1<-cbind(weekday.steps.mean$steps,weekday.steps.mean$interval)
d3<-data.frame(d1,WeekDay="weekday")
dat2<-rbind(d2,d3)
names(dat2)<-c("MeanSteps","interval", "Day")
dat2<-dat2[order(as.numeric(dat2$interval)),]
panel.means <- function(x, y) {
        panel.xyplot(x, y, type="l")
        panel.abline(h=mean(y),
                     col="red")
        }
scatter<-xyplot(as.numeric(MeanSteps)~as.integer(interval)|as.factor(Day),
       data=dat2,
       ylab="Mean Number Daily Steps",
       main="Activity Patterns: Weekdays and Weekends",
       panel=panel.means)
print(scatter)
trellis.device(device="png", filename="plot1.png")
print(scatter)
dev.off()
```

The activity pattern shows distintively different patterns of activity for the weekend and the weekdays.  The mean number of steps for weekday (`r mean.steps.weekday`) and weekend (`r mean.steps.weekend`) (the red lines on the "Activity Patterns" plot show the difference in mean activity).



```{r boxPlots, fig.align='center'}
bwplot(as.numeric(MeanSteps)~as.factor(Day),
       data=dat2,
       ylab="Mean Number Daily Steps",
       par.settings = list( box.umbrella=list(col= c("orange", "blue")),
                            box.dot=list(col= c("orange", "blue")),
                            box.rectangle = list(col= c("orange", "blue"))),
       main="Activity Patterns: Weekdays and Weekends")
```

